Fast Sorting on Modern Computers

### 2005-08-01 docs/tech_report/quick.tex
more minor renaming
enambled main again
more minor edits.

An additional benefit of Sedgewick's median of three implementation is that it
provides a sentinel for the partition. It is now possible to remove the bounds
check from the innermost loops of \cc{partition}, saving a lot of instructions.

@@

Quicksort has been well studied, and several important improvements have been
noted. In \cite{BentleyMcIlroy93}, using a median of nine partition is discussed,
and modern discussions on quicksort refer to the problem of equal keys, which is
discussed there, as well as by Sedgewick in \cite{SedgewickBentley02}. These
improvements are not included here for several reasons. Firstly, a quicksort
which takes into account equal keys puts the equal keys at the front of the
array, then copies them into place later. These copies would represent a lot of
cache misses, and the results would overwhelm the improvements LaMarca makes in
the next phase of the algorithm. Secondly, when sorting random keys over a
large range, the number of equal keys will be very low. Using a median of
greater than three is discussed in Section \ref{quick-predictors2}.

\section{Memory-Tuned Quicksort}

@@

The instruction count for each quicksort is expected to be $O(NlogN)$, though
not as high as the other algorithms due to the efficient inner loop.  The number
of instructions for multi-quicksort should increase by about 50\% in the case of
the sequential sort, and 25\% in the case of the binary sort. Cache quicksort
should also have a slightly higher instruction count than base quicksort, due to
the change in use of insertion sort.

@@

The results show again that direct mapped caches can perform better than fully
associative ones. This is true in the case of all the quicksorts, as can be seen
in Figures \ref{base quicksort (median-of-3) (no median) level 2 misses} to
\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
in the case of all the quicksorts.

@@

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base,
cache and multi quicksorts and base quicksort (pseudo-median-of-9)}

@@

\label{quick-predictors2}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}

@@

We use several variations of quicksort to test this theory. \ref{bently, i
ithkn} suggests partitioning using a median-of-9 instead of a median-of-3. We
created sorts using no median, a median-of-3, and pseudo-medians-of-five, -seven and
-nine. The pseudo-medians used repeated medians-of-three instead of a real
median, as suggestion by \ref{bently, again, i think}. 
%use a cut off

@@

The results of these tests are shown in Figure \ref{Branch prediction
performance for base quicksort without a median, with median of 3, and with
pseudo medians of 5, 7 and 9}. There is little variance in insertion sort misses
between the sorts using pseudo-medians, but the median-of-three quicksort has
6\% fewer branches, and a corresponsing decrease in mispredictions.
%TODO to whom do we owe the idea of a pseudo median? Bently/McIlroy?
%TODO when does it cut out, and why did we choose tere (benlty?)

Between a median-of-three and a pseudo-median-of-nine partitioning, there is a
10\% reduction in branches. At the same time, there is a 20\% reduction in
correct predictions, accounting for the entire reduction in branches, and a 4\%
increase in mispredictions. This is due to better partitioning. Less time is
spent copying keys unnecessarily, resulting in a less predictable branch. The
corresponding increase in mispredictions while calculating the median equates to
50\% of the extra mispredictions due to partitioning.

@@

The pseudo-median-of-5 partitioning performs slightly better than the
median-of-3, and pseudo-median-of-7 performs slightly better again, in
terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
the quicksorts, showing that a pseudo-median-of-seven offers the greatest
trade-off between the reduced number of branches and instructions, and the
increase in branch mispredictions.

>>>>

An additional benefit of Sedgewick's median-of-3 implementation is that it
provides a sentinel for the partition. It is now possible to remove the bounds
check from the innermost loops of \cc{partition}, saving a lot of instructions.

@@

Quicksort has been well studied, and several important improvements have been
noted. In \cite{BentleyMcIlroy93}, using a median-of-9 partition is discussed,
and modern discussions on quicksort refer to the problem of equal keys, which is
discussed there, as well as by Sedgewick in \cite{SedgewickBentley02}. These
improvements are not included here for several reasons. Firstly, a quicksort
which takes into account equal keys puts the equal keys at the front of the
array, then copies them into place later. These copies would represent a lot of
cache misses, and the results would overwhelm the improvements LaMarca makes in
the next phase of the algorithm. Secondly, when sorting random keys over a
large range, the number of equal keys will be very low. Using a median of
greater than three is discussed in Section \ref{quick-predictors2}.

\section{Memory-tuned Quicksort}

@@

The instruction count for each quicksort is expected to be $O(NlogN)$, though
not as high as the other algorithms due to the efficient inner loop.  The number
of instructions for multi-quicksort should increase by about 50\% in the case of
the sequential sort, and 25\% in the case of the binary sort. Memory-tuned
quicksort should also have a slightly higher instruction count than base
quicksort, due to the change in use of insertion sort.

@@

The results show again that direct mapped caches can perform better than fully
associative ones. This is true in the case of all the quicksorts, as can be seen
in Figures \ref{base quicksort (no median) level 2 misses} to
\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
in the case of all the quicksorts.

@@

\sixplot{base_quicksort__median_of_3__0.eps}{base_quicksort__pseudo_median_of_9__0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort__binary_search__0.eps}{multi_quicksort__sequential_search__0.eps}{}{base,
cache and multi quicksorts and base quicksort (pseudo-median-of-9)}

@@

\label{quick-predictors2}
\sixplot{base_quicksort__no_median__0.eps}{base_quicksort__median_of_3__0.eps}{base_quicksort__pseudo_median_of_5__0.eps}{base_quicksort__pseudo_median_of_7__0.eps}{base_quicksort__pseudo_median_of_9__0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}

@@

We use several variations of quicksort to test this theory. \ref{bently, i
ithkn} suggests partitioning using a median-of-9 instead of a median-of-3. We
created sorts using no median, a median-of-3, and pseudo-medians-of-5, -7 and
-9. The pseudo-medians used repeated medians-of-3 instead of a real
median, as suggestion by \ref{bently, again, i think}. 
%use a cut off

@@

The results of these tests are shown in Figure \ref{Branch prediction
performance for base quicksort without a median, with median of 3, and with
pseudo medians of 5, 7 and 9}. There is little variance in insertion sort misses
between the sorts using pseudo-medians, but the median-of-3 quicksort has
6\% fewer branches, and a corresponsing decrease in mispredictions.
%TODO to whom do we owe the idea of a pseudo median? Bently/McIlroy?
%TODO when does it cut out, and why did we choose tere (benlty?)

Between a median-of-3 and a pseudo-median-of-9 partitioning, there is a
10\% reduction in branches. At the same time, there is a 20\% reduction in
correct predictions, accounting for the entire reduction in branches, and a 4\%
increase in mispredictions. This is due to better partitioning. Less time is
spent copying keys unnecessarily, resulting in a less predictable branch. The
corresponding increase in mispredictions while calculating the median equates to
50\% of the extra mispredictions due to partitioning.

@@

The pseudo-median-of-5 partitioning performs slightly better than the
median-of-3, and pseudo-median-of-7 performs slightly better again, in
terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
the quicksorts, showing that a pseudo-median-of-7 offers the greatest
trade-off between the reduced number of branches and instructions, and the
increase in branch mispredictions.
